RTS AI: Problems and Techniques

### 2015-01-19 ECGG_RTSProblems/ECGG_RTSProblems.tex
smaller corrections/additions

\section{Introduction}\label{sec:intro}
The  field of  Real-Time Strategy  (RTS) game  Artificial Intelligence
(AI) has advanced significantly in the past few year, partially thanks to competitions like the ``ORTS
RTS  Game AI  Competition''  (held  from 2006  to  2009), the  ``AIIDE
StarCraft AI Competition'' (held since  2010), and the ``CIG StarCraft
RTS AI Competition'' (held since  2011). Based on the work work presented in \cite{ontanon2013survey}, here we first define RTS games, and then list here the open problems in creating AI for RTS games, and the approaches that have been proposed to address these problems.

@@

\subsubsection{Uncertainty}
Adversarial planning under  uncertainty in domains of the  size of RTS
games is  still an unsolved  challenge.  In  RTS games, there  are two
main kinds  of uncertainty. First,  the game is  partially observable,
and players  cannot observe the  whole game  map (like in  Chess), but
need to scout in order to see what the opponent is doing. This type of
uncertainty   can  be   lowered  by   good  scouting,   and  knowledge
representation  (to  infer  what  is  possible  given  what  has  been
seen). Second,  there is also  uncertainty arising from the  fact that
the games  are adversarial,  and a player  cannot predict  the actions
that the opponent(s)  will execute. For this type  of uncertainty, the
AI, as the human  player, can only build a sensible  model of what the
opponent is likely to do.
% [Santi:  I  commented  out  this  part,  since  I'm  not  sure  it's
% correct.  The second  type  of  uncertainty can  also  be linked  to
% data.  For example,  by analyzing  player replays,  one could  learn
% "plan recognition"  probability tables, which  can be used  to infer
% the strategy  of the opponent,  and thus, reduce the  uncertainty of
% the opponent's future  actions] Both these kinds  of uncertainty can
% be studied through probabilities, but only the first type, $P(Hidden
% | Seen)$, can be {\em easily} linked to
% data. %now for the clumsy part:
% One could  argue that the  extensional data  is the only  thing that
% matters because  it is the only  thing that has an  influence on the
% game, but in  reality, for very highly  skilled players, discovering
% the intentions of the  enemy is the key to winning.  It allows for a
% compact understanding of the game  state and development, while also
% giving additional information about the  future. Whereas a player is
% doing action  A to  follow it  up by action  B or  C depends  on its
% intention,  while the  information  about action  A  is visible  and
% information about  B or  C is  hidden, only  intention allows  for a
% choice.

@@

Systems that  play RTS  games need  to address most,  if not  all, the
aforementioned problems  together. Therefore,  it is hard  to classify
existing  work  on  RTS  AI   as  addressing  the  different  problems
above. For that reason, %in order to classify existing work on RTS AI,
we will divide  it according to three levels  of abstraction: strategy
(which loosely corresponds to ``macro''), tactics and reactive control
(which loosely corresponds to ``micro'').

Figure~\ref{fig:levels-abstraction}   graphically    illustrates   how
strategy, tactics and reactive control are three points in a continuum
scale where  strategy corresponds  to decisions making  processes that
affect long spans of time (several  minutes in the case of StarCraft),
reactive control corresponds  to low-level second-by-second decisions,
and tactics sit in the  middle. Also, strategic decisions reason about
the whole game at once, whereas tactical or reactive control decisions
are localized,  and affect only  specific groups of  units. Typically,
strategic decisions constrain future tactical decisions, which in turn
condition  reactive  control.  Moreover,  information  gathered  while
performing reactive control, can  cause reconsideration of the tactics
being employed; which could trigger further strategic reasoning.

@@

% PLANNING:
\subsubsection{Planning}
Approaches using  planning techniques have  also been explored  in the
literature.  For example  Onta\~{n}\'{o}n  et al.  \cite{CBR_Planning}
explored the use of real-time  case-based planning (CBP) in the domain
of  Wargus (a  Warcraft  II clone).  In their  work,  they used  human
demonstration to learn  plans, which are then composed  at run-time in
order  to   form  full-fledges  strategies   to  play  the   game.  In
\cite{PlanRetrieval} they improve over  their previous CBP approach by
using situation assessment for improving the quality and speed of plan
retrieval.  Hierarchical Task-Network  (HTN)  planning  has also  been
explored  with some  success in  the context  of simpler  first-person
shooter  games  \cite{HTNPlanning}.  Planning  approaches  offer  more
adaptivity    of   the    AI   strategy    compared   to    hard-coded
approaches. However, the real-time constraints  of RTS games limit the
planning approaches that  can be applied, HTN  and case-based planning
being  the  only  ones  explored  so  far.  Moreover,  none  of  these
approaches addresses any timing or scheduling issues, which are key in
RTS games.  On notable  exception is  the work  of Churchill  and Buro
\cite{churchill2011build}, who used planning in order to construct its
economic build-orders,  taking into account timing  constraints of the
different actions.

@@

% POTENTIAL FIELDS:
\subsubsection{Potential  fields}
Potential  fields and  influence maps  have  been found  to be  useful
techniques for reactive decision making. Some uses of potential fields
in RTS  games are: avoiding obstacles  (navigation), avoiding opponent
fire \cite{uriarte2012kiting}, or staying at maximum shooting distance
\cite{Hagelback09}. Potential  fields have also been  combined with A*
path-finding  to avoid  local traps  \cite{Hagelback12}. Hagelb\"{a}ck
and  Johansson \cite{HagelbackJ08}  presented a  multi-agent potential
fields  based bot  able  to  deal with  fog-of-war  in the  Tankbattle
game. Avery et  al. \cite{Avery09} and Smith  et al. \cite{SmithCIG10}
co-evolved  influence   map  trees   for  spatial  reasoning   in  RTS
games. Danielsiek et al. \cite{Danielsiek_2008} used influence maps to
achieve  intelligent squad  movement to  flank the  opponent in  a RTS
game.  Despite their  success,  a drawback  for potential  field-based
techniques is the  large number of parameters that has  to be tuned in
order to  achieve the  desired behavior. Approaches  for automatically
learning  such  parameters  have  been explored,  for  example,  using
reinforcement learning \cite{Liu_2008},  or self-organizing-maps (SOM)
\cite{teamCompositionRTS}. We would like to note that potential fields
are a reactive control technique, and as such, they do not perform any
form of  lookahead. As  a consequence, these  techniques are  prone to
make units stuck in local maxima.

>>>>

\section{Introduction}\label{sec:intro}
The  field of  Real-Time Strategy  (RTS) game  Artificial Intelligence
(AI) has advanced significantly in the past few years, partially thanks to competitions as the ``ORTS
RTS  Game AI  Competition''  (held  from 2006  to  2009), the  ``AIIDE
StarCraft AI Competition'' (held since  2010), and the ``CIG StarCraft
RTS AI Competition'' (held since  2011). Based on the work work presented in \cite{ontanon2013survey}, here we first define RTS games, then list the open problems in creating AI for RTS games, and finally point to the approaches that have been proposed to address these problems.

@@

\subsubsection{Uncertainty}
Adversarial planning under  uncertainty in domains of the  size of RTS
games is  still an unsolved  challenge.  In  RTS games, there  are two
main kinds  of uncertainty. First,  the game is  partially observable,
and players  cannot observe the  whole game  map (like in  Chess), but
need to scout in order to see what the opponent is doing. This type of
uncertainty   can  be   lowered  by   good  scouting,   and  knowledge
representation  (to  infer  what  is  possible  given  what  has  been
seen). However, scouting also means deliberately reducing economic progress in order to obtain information.
Second,  there is also  uncertainty arising from the  fact that
the games  are adversarial,  and a player  cannot predict  the actions
that the opponent(s)  will execute. For this type  of uncertainty, the
AI, as the human  player, can only build a sensible  model of what the
opponent is likely to do.
% [Santi:  I  commented  out  this  part,  since  I'm  not  sure  it's
% correct.  The second  type  of  uncertainty can  also  be linked  to
% data.  For example,  by analyzing  player replays,  one could  learn
% "plan recognition"  probability tables, which  can be used  to infer
% the strategy  of the opponent,  and thus, reduce the  uncertainty of
% the opponent's future  actions] Both these kinds  of uncertainty can
% be studied through probabilities, but only the first type, $P(Hidden
% | Seen)$, can be {\em easily} linked to
% data. %now for the clumsy part:
% One could  argue that the  extensional data  is the only  thing that
% matters because  it is the only  thing that has an  influence on the
% game, but in  reality, for very highly  skilled players, discovering
% the intentions of the  enemy is the key to winning.  It allows for a
% compact understanding of the game  state and development, while also
% giving additional information about the  future. Whereas a player is
% doing action  A to  follow it  up by action  B or  C depends  on its
% intention,  while the  information  about action  A  is visible  and
% information about  B or  C is  hidden, only  intention allows  for a
% choice.

@@

Systems that  play RTS  games need  to address most,  if not  all, the
aforementioned problems  together. Therefore,  it is hard  to classify
existing  work  on  RTS  AI   as  addressing  the  different  problems
above. For that reason, %in order to classify existing work on RTS AI,
we will divide  it according to three levels  of abstraction: strategy
(which loosely corresponds to ``macro''), tactics, and reactive control
(which loosely corresponds to ``micro'').

Figure~\ref{fig:levels-abstraction}   illustrates   how
strategy, tactics, and reactive control are three points in a continuum
scale where  strategy corresponds  to decisions making  processes that
affect long spans of time (several  minutes in the case of StarCraft),
reactive control corresponds  to low-level second-by-second decisions,
and tactics sit in the  middle. Also, strategic decisions reason about
the whole game at once, whereas tactical or reactive control decisions
are localized,  and affect only  specific groups of  units. Typically,
strategic decisions constrain future tactical decisions, which in turn
condition  reactive  control.  Moreover,  information  gathered  while
performing reactive control, can  cause reconsideration of the tactics
being employed; which could trigger further strategic reasoning.

@@

% PLANNING:
\subsubsection{Planning}
Approaches using  planning techniques have  also been explored  in the
literature.  For example  Onta\~{n}\'{o}n  et al.  \cite{CBR_Planning}
explored the use of real-time  case-based planning (CBP) in the domain
of  Wargus (a  Warcraft  II clone).  In their  work,  they used  human
demonstration to learn  plans, which are then composed  at run-time in
order  to   form  full-fledged  strategies   to  play  the   game.  In
\cite{PlanRetrieval} they improve over  their previous CBP approach by
using situation assessment for improving the quality and speed of plan
retrieval.  Hierarchical Task-Network  (HTN)  planning  has also  been
explored  with some  success in  the context  of simpler  first-person
shooter  games  \cite{HTNPlanning}.  Planning  approaches  offer  more
adaptivity    of   the    AI   strategy    compared   to    hard-coded
approaches. However, the real-time constraints  of RTS games limit the
planning approaches that  can be applied, HTN  and case-based planning
being  the  only  ones  explored  so  far.  Moreover,  none  of  these
approaches addresses any timing or scheduling issues, which are key in
RTS games.  On notable  exception is  the work  of Churchill  and Buro
\cite{churchill2011build}, who used planning in order to construct its
economic build-orders,  taking into account timing  constraints of the
different actions.

@@

% POTENTIAL FIELDS:
\subsubsection{Potential  fields}
Potential  fields and  influence maps  have  been found  to be  useful
techniques for reactive decision making. Some uses of potential fields
in RTS  games are: avoiding obstacles  (navigation), avoiding opponent
fire \cite{uriarte2012kiting}, or staying at maximum shooting distance
\cite{Hagelback09}. Potential  fields have also been  combined with A*
path-finding  to avoid  local traps  \cite{Hagelback12}. Hagelb\"{a}ck
and  Johansson \cite{HagelbackJ08}  presented a  multi-agent potential
fields  based bot  able  to  deal with  fog-of-war  in the  Tankbattle
game. Avery et  al. \cite{Avery09} and Smith  et al. \cite{SmithCIG10}
co-evolved  influence   map  trees   for  spatial  reasoning   in  RTS
games. Danielsiek et al. \cite{Danielsiek_2008} used influence maps to
achieve  intelligent squad  movement to  flank the  opponent in  a RTS
game.  Despite their  success,  a drawback  for potential  field-based
techniques is the  large number of parameters that has  to be tuned in
order to  achieve the  desired behavior. Approaches  for automatically
learning  such  parameters  have  been explored,  for  example,  using
reinforcement learning \cite{Liu_2008},  or self-organizing-maps (SOM)
\cite{teamCompositionRTS}. We would like to note that potential fields
are a reactive control technique, and as such, they do not perform any
form of  lookahead. As  a consequence, these  techniques are  prone to
make units stuck in local optima.
